Left-linear rewriting in adhesive categories

### 2024-04-23 concur/fusione.tex
minima

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for so-called linear rules which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only for the order of
independent steps.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify quite general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we also prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely

@@

    A $\mor(\X)$-Van Kampen square will be called simply \emph{Van
      Kampen} and similarly, a $\mor(\X)$-stable square is called
    \emph{stable}.
  \end{definition}

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor categories \cite{mac2013categories}, as shown by the following theorems from~\cite{ehrig2006fundamentals,lack2005adhesive}.

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-behaved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (e.g., for graphs, nodes can be
merged while edges cannot). The idea here is to consider graphical structures as 
presheaves on categories in which there are objects which play the role of 
\emph{roots}, i.e.~objects which are not codomains of any arrow besides 
the identity. \todo{D: quest'ultima frase ha senso? A: Ok. Ho cambiato ``as graphs'' con ``graphical structures''.}%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$. The
	category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that  the
	following hold for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$, the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

@@

For instance, the category $\cat{Graph}$ can be obtained
by taking as $\X$ the category generated by $E \rightrightarrows V$. In this case $E$ is the only root, hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is mono on edges, i.e.~it can only merge nodes.

@@

Left-linear rewriting systems on $\cat{EGraphs}$,\todo{A: sostituirei ``, i.e. in which'' con ``where''. Giusto?} i.e.~in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\begin{example}
  \label{ex:non-global}
  Consider a new rewriting system in $\cat{Graph}$ consisting of the
  rules in $\rho_0$, $\rho_1$ and $\rho_2$ in
  Fig.~\ref{fi:rules-cas}.\todo{P: rev.2 dice ``In the top-left and
    top-middle nodes of $\rho_3$, the labels "12" should be "1" and
    "2".''. Non e' corretto, ma magari si capisce poco ...}  Consider
  the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\rho_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\rho_0$ or by $\rho_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\rho_2$. Indeed, in $\der{F}$ it can be easily seen that the second
  and the third step applying $\rho_1$ and $\rho_2$ are
  independent. However, we could switch twice the application of
  $\rho_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\rho_1$ and $\rho_2$ are no longer independent.

@@

The lemmas above open the way to the key result of this section establishing a canonical form for
switching derivations.

The first result shows that whenever equating two derivation sequences by performing a sequence of switching, we can limit ourselves to switchings inverting the order of steps that in the target derivation has to be reversed, i.e.~we can limit ourselves to applying inversions.

@@

While inversions can be
performed in any order in the case of linear rewriting systems, this is no longer true for left-linear
rewriting systems due to the fact that, as already observed, globality of independence holds in a weak form. For instance, if we get back to Example~\ref{ex:non-global} and consider derivation $\der{F}$ in Fig.~\ref{fi:derF}, the last two steps applying rules $\rho_1$ and $\rho_2$ can be switched, thus leading to a derivation, call it $\der{F}''$,  applying $\rho_0, \rho_2, \rho_1$. Such $\der{F}''$ is switch equivalent to $\der{F}'$ in Fig.~\ref{fi:derF1}, hence starting from $\der{F}''$ we can obtain $\der{F}'$ via a switching sequence, but we cannot start by switching $\rho_0$ and $\rho_2$, despite the fact that this is an inversion.

@@

The class of well-switching $\mathcal{M}$-adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of $\mathcal{M}$-adhesive categories that include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the sorts 
of depth 0  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

@@

On the practical side, the present paper just surveyed the possibility
of modelling e-graphs. This appears to be quite interesting as
concurrent rewriting on such structures is an active area of
research~\cite{abs-2208-06295}. \todo{F: altre citazioni?} 
We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on e-graphs.

>>>>

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules some fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only for the order of
independent steps.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify quite general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely

@@

    A $\mor(\X)$-Van Kampen square will be called simply \emph{Van
      Kampen} and similarly a $\mor(\X)$-stable square will be called
    \emph{stable}.
  \end{definition}

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor categories \cite{mac2013categories}, as shown by the following theorems~\cite{ehrig2006fundamentals,lack2005adhesive}.

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but this is expected, since natural examples are all in the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (for graphs, nodes can be
merged while edges cannot). The idea here is to consider as graphs 
those presheaves on categories in which there are object that plays 
the roles of \emph{roots}, i.e.~objects that are not the codomain of 
any arrow besides the identity). 
%\todo{D: quest'ultima frase ha senso?}
%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$.
	The category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$ hold
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$ the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

@@

For instance, the category $\cat{Graph}$ can be obtained
by taking as $\X$ the category generated by $E \rightrightarrows V$. In this case $E$ is the only root, hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is injective on edges, i.e.~it can only merge nodes.

@@

Left-linear rewriting systems on $\cat{EGraphs}$, i.e.~in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\begin{example}
  \label{ex:non-global}
  Consider a new rewriting system in $\cat{Graph}$ consisting of the
  rules in $\sigma_0$, $\sigma_1$, and $\sigma_2$ in
  Fig.~\ref{fi:rules-cas}.
  %\todo{P: rev.2 dice ``In the top-left and
   % top-middle nodes of $\rho_3$, the labels "12" should be "1" and
   % "2".''. Non e' corretto, ma magari si capisce poco ...}  
   Consider the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\sigma_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\sigma_0$ or by $\sigma_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\sigma_2$. Indeed, in $\der{F}$ it can be easily seen that the second
  and the third step applying $\sigma_1$ and $\sigma_2$ are
  independent. However, we could switch twice the application of
  $\sigma_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\sigma_1$ and $\sigma_2$ are no longer independent.

@@

The lemmas above open the way to the key result of this section, which establishes a canonical form for
switching derivations.

The first result shows that when equating two derivation sequences by performing a series of switchings, we can limit ourselves to switchings that inverte the order of steps that in the target derivation has to be reversed, i.e.~we can limit ourselves to applying inversions.

@@

While inversions can be
performed in any order in the case of linear rewriting systems, this is no longer true for left-linear
rewriting systems due to the fact that, as already observed, globality of independence holds in a weak form. 
For instance, if we get back to Example~\ref{ex:non-global} and consider derivation $\der{F}$ in Fig.~\ref{fi:derF}, 
the last two steps applying rules $\sigma_1$ and $\sigma_2$ can be switched, thus leading to a derivation, 
call it $\der{F}''$,  applying $\sigma_0, \sigma_2, \sigma_1$. Such $\der{F}''$ is switch equivalent to $\der{F}'$ in Fig.~\ref{fi:derF1}, 
hence starting from $\der{F}''$ we can obtain $\der{F}'$ via a switching sequence, but we cannot start 
by switching $\sigma_0$ and $\sigma_2$, despite the fact that this is an inversion.

@@

The class of well-switching $\mathcal{M}$-adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of $\mathcal{M}$-adhesive categories that include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements belonging to the sorts 
of the roots  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

@@

On the practical side, the present paper just surveyed the possibility
of modelling e-graphs. This appears to be quite interesting as
concurrent rewriting on such structures is an active area of
research~\cite{abs-2208-06295}. 
%\todo{F: altre citazioni?} 
We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on e-graphs.
